### 2017-08-10 correctnesshpc/paper.tex
Added initial paper and fixed let semantics

\title{A Formal Specification of Legion}
\author{George Stelle \inst{1} \and Noah Evans \inst{2}}
\institute{Los Alamos National Laboratory \\ 
\email{stelleg@lanl.gov}
\and Sandia National Laboratories 
\email{nevans@sandia.gov}}

\begin{abstract}
Legion is a programming model that constrains effects in a way that ensures
serial semantics are preserved. We formalize the Legion semantics in a
mechanized proof assistant, Coq, and show how this enables one to reason
formally and prove theorems about the semantics. We also discuss how to extend
this work to prove that serial semantics are preserved and how given enough
time, one could extend the effort to prove an implementation is correct.
\end{abstract}

\section{Intro}
High performance computing, as the name implies, has long focused on
performance, and for good reason. Expensive machines are purchased to run
increasingly complex simulations, and every optimization can save huge amounts
of money, both due to the initial cost of the machine and the cost of power to
run it.

That said, what good is a computation if we aren't relatively certain that the
answers it returns are correct? Presumably, these computations inform important
decisions about topics ranging from climate to nuclear safety. We should be as
confident as we can be that we aren't basing these decisions on bugs.

\section{Background}

Legion is a programming model for high performance computing that enables
reasoning about with the attractive property that it is
guaranteed to preserve serial semantics. It achieves this by dividing memory
into logical regions. This allows programmers to avoid the
burden

\subsection{Theorem Provers}

Improvements to proof assistants over recent decades have made them a powerful
tool for implementing and reasoning about code. From provably correct compilers
to provably correct operating systems, proof assistants enable extremely high
certainty in correctness even for large scale software projects.

\section{A Formal Specification of Legion}

"A program without a specification cannot be incorrect, it can only be
surprising." TODO: Find quote source in recent Appel paper. One of the most
important and challenging aspects of proving correctness is defining a formal
specification of what a program *should* do. Indeed, there is currently a large
NSF effort dedicated to improving our ability to do just this \cite{deepspec}.

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{oopsla13}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it.

\section{Discussion}

One common theme in formalizing systems is the discovery of holes in the
informal specification. For example, when formalizing the C language for the
CompCert project, Leroy et. al. found many cases where the specification was
incomplete. Note that this is seperate notion from undefined behavior. Undefined
behavior in a specification is well defined. A little bit like known unknowns
vs. unknown unknowns, gaps in a specification

In formalizing Legion, we discovered that while there is an impressive theory
ensuring that the side effects of function calls running concurrently preserve
serial semantics, there is no theory for what function calls can run
concurrently based on value dependencies. For example, consider the infamous
fibonacci sequence example:

\begin{lstlisting}
fib(n) = if n < 2 then 1 else fib(n-1) + fib(n-2)
\end{lstlisting}

It is well known that the function calls require

### 2017-08-11 correctnesshpc/paper.tex
Small changes to paper

@@
-\title{A Formal Specification of Legion}
+\title{Towards Correct Legion}
 \author{George Stelle \inst{1} \and Noah Evans \inst{2}}
@@
 \begin{abstract}
-Legion is a programming model that constrains effects in a way that ensures
-serial semantics are preserved. We formalize the Legion semantics in a
-mechanized proof assistant, Coq, and show how this enables one to reason
-formally and prove theorems about the semantics. We also discuss how to extend
-this work to prove that serial semantics are preserved and how given enough
-time, one could extend the effort to prove an implementation is correct.
+Legion is a programming model that uses logical regions of memory to reason about 
+locality and concurrency. We formalize the Legion semantics in a mechanized
+proof assistant, Coq, and show how this enables one to reason formally and
+prove theorems about the semantics. We also discuss how to extend this work to
+prove that serial semantics are preserved and how given enough time, one could
+extend the effort to prove an implementation is correct.
 \end{abstract}
@@
 
-That said, what good is a computation if we aren't relatively certain that the
-answers it returns are correct? Presumably, these computations inform important
-decisions about topics ranging from climate to nuclear safety. We should be as
-confident as we can be that we aren't basing these decisions on bugs.
+That said, what good is a fast computation if we aren't relatively certain that
+the answers it returns are correct? Presumably, these computations inform
+important decisions about topics ranging from climate to nuclear safety. We
+should be as confident as possible that we aren't basing these decisions on
+unexpected semantics.
 
@@
 Legion is a programming model for high performance computing that enables
-reasoning about with the attractive property that it is
-guaranteed to preserve serial semantics. It achieves this by dividing memory
-into logical regions. This allows programmers to avoid the
+reasoning about locality and with the attractive property that it is guaranteed
+to preserve serial semantics. It achieves this by dividing memory into logical
+regions.  This allows programmers to avoid the
 burden
-
-\subsection{Theorem Provers}
 
@@
 certainty in correctness even for large scale software projects.
+
+Coq is a dependently typed proof assistant. At its core is the Calculus of
+Inductive Constructions, a small dependently typed programming language and
+logic, capable of powerful abstractions for formal proofs. In the last 10 years 
+great strides have been made in using it to both build systems and prove useful
+properties about them \cite{compcert, certicoq, chlipala...}.
+
+Specifications come in a wide range of formats. Many, like the OpenMP
+specification, are very informal, primarily consisting of english. Others, like
+Legion's previously, are written by hand in the using standard notation for
+logical relations, etc. For this paper, when we refer to a \emph{formal}
+specification, we are referring to a specification defined in a formal logic,
+like Coq.
+
+While semantics defined in papers are a huge improvement over natural language
+specifications, there is still room for improvement. \cite{formalspec vs paper} has 
+shown that even mathematical specifications and proofs in computer science
+papers almost always have errors that would be caught by a formal proof. This is not to say,
+of course, that formal specifications and proofs are infallible. On the contrary, it 
+is quite easy to get a specification wrong. That said, often, if a there is a mistake in a spec
+, the process of proving it formally will expose that mistake, allowing for its correction.
 
@@
 One common theme in formalizing systems is the discovery of holes in the
-informal specification. For example, when formalizing the C language for the
-CompCert project, Leroy et. al. found many cases where the specification was
-incomplete. Note that this is seperate notion from undefined behavior. Undefined
-behavior in a specification is well defined. A little bit like known unknowns
-vs. unknown unknowns, gaps in a specification
+corresponding informal specification. For example, when formalizing the C
+language for the CompCert project, Leroy et. al. found many cases where the
+specification was incomplete. Note that this is seperate notion from undefined
+behavior. Undefined behavior in a specification is well defined. A little bit
+like known unknowns vs. unknown unknowns, gaps in a specification
 
@@
 concurrently based on value dependencies. For example, consider the infamous
-fibonacci sequence example:
+Fibonacci example, which naively computes the \texttt{n}'th Fibonacci number:
 
@@
 
-It is well known that the function calls require
+Of course

### 2017-08-15 correctnesshpc/paper.tex
Fixed up formatting
Added conclusion and other text

@@
+\documentclass[sigplan]{acmart}
+
+\usepackage{booktabs} % For formal tables
+\usepackage{listings} % For formal tables
+
+% Copyright
+%\setcopyright{none}
+%\setcopyright{acmcopyright}
+%\setcopyright{acmlicensed}
+\settopmatter{printacmref=false}
+\setcopyright{rightsretained}
+%\setcopyright{usgov}
+%\setcopyright{usgovmixed}
+%\setcopyright{cagov}
+%\setcopyright{cagovmixed}
+
+% DOI
+\acmDOI{10.475/123_4}
+
+% ISBN
+\acmISBN{123-4567-24-567/08/06}
+
+%Conference
+\acmConference[Correctness'17]{Correctness 2017: First International Workshop on
+  Software Correctness for HPC Applications}{November 2017}{Denver Colorado, USA} 
+\acmYear{2017}
+\copyrightyear{2017}
+
+%\acmPrice{15.00}
+
+\begin{document}
+
 \title{Towards Correct Legion}
-\author{George Stelle \inst{1} \and Noah Evans \inst{2}}
-\institute{Los Alamos National Laboratory \\ 
+\author{George Stelle}
+\affiliation{\institution{Los Alamos National Laboratories}}
 \email{stelleg@lanl.gov}
-\and Sandia National Laboratories 
-\email{nevans@sandia.gov}}
+
+\author{Noah Evans}
+\affiliation{\institution{Sandia National Laboratories}}
+\email{nevans@sandia.gov}
 
@@
 
+\maketitle
+
 \section{Intro}
 High performance computing, as the name implies, has long focused on
-performance, and for good reason. Expensive machines are purchased to run
-increasingly complex simulations, and every optimization can save huge amounts
-of money, both due to the initial cost of the machine and the cost of power to
-run it.
+performance. Expensive machines are purchased to run increasingly complex
+simulations, and every optimization can save huge amounts of money, both due to
+the initial cost of the machine and the cost of power to run it.
 
-That said, what good is a fast computation if we aren't relatively certain that
-the answers it returns are correct? Presumably, these computations inform
-important decisions about topics ranging from climate to nuclear safety. We
-should be as confident as possible that we aren't basing these decisions on
-unexpected semantics.
+That said, what good is a fast computation if we aren't certain that the
+values it computes are correct? Presumably, these computations inform important
+decisions about topics ranging from climate to nuclear safety. We should be as
+confident as possible that we aren't basing these decisions on unexpected
+semantics.
+
+Recent work has shown that modular, composable formal specifications enable
+formal correctness proofs for large software systems that were previously
+unattainable. By integrating a formal specification tightly with an
+implementation, one can achieve the highest levels of confidence in code
+currently possible \cite{?}. For example, the Compcert project implements an
+optimizing c compiler and proves it correct. The success of Compcert in
+removing compiler bugs has been well documented. We will discuss later how 
+Compcert could be used in building verifiable HPC systems.
+
+In HPC, programming models have become an increasingly important way to ensure
+that code stays maintanable, performant, and portable. We argue that these
+programming models provide a good target for formalization: any correctness
+arguments can be leveraged by every program using the model. It is exactly this 
+kind of modularization of proofs that has been so successful in proving programs 
+correct \cite{?}.
+
+For this paper, we take a recent programming model, Legion, and formalize its
+semantics and type system. We then prove a few supporting lemmas as examples of
+the kinds of things one can prove in an HPC settings. Finally, we discuss the
+some of the possible paths towards full, formally correct implementation. This
+is no small task, as we discuss later, and the ways forward are numerous and
+lengthy. But if we believe that HPC informs important decisions, we must strive
+for this pinnacle of software correctness!
 
@@
 specifications, there is still room for improvement. \cite{formalspec vs paper} has 
-shown that even mathematical specifications and proofs in computer science
-papers almost always have errors that would be caught by a formal proof. This is not to say,
-of course, that formal specifications and proofs are infallible. On the contrary, it 
-is quite easy to get a specification wrong. That said, often, if a there is a mistake in a spec
-, the process of proving it formally will expose that mistake, allowing for its correction.
+shown that even hand-written mathematical specifications and proofs in computer
+science papers almost always have errors that can be caught by a formal proof.
+This is not to say, of course, that formal specifications and proofs are
+infallible.  On the contrary, it is quite easy to get a specification wrong.
+That said, often, if a there is a mistake in a spec , the process of proving it
+formally will expose that mistake, allowing for its correction.
 
@@
 "A program without a specification cannot be incorrect, it can only be
-surprising." TODO: Find quote source in recent Appel paper. One of the most
-important and challenging aspects of proving correctness is defining a formal
-specification of what a program *should* do. Indeed, there is currently a large
-NSF effort dedicated to improving our ability to do just this \cite{deepspec}.
+surprising." \cite{appel}. One of the most important and challenging aspects of
+proving correctness is defining a formal specification of what a program
+*should* do. Indeed, there is currently a large NSF effort dedicated to
+improving our ability to do just this \cite{deepspec}. The choice of
+specification has far-reaching consequences in how easy it is to later prove
+correctness properties. For example, if a specification is chosen poorly,
 
@@
 We take this specification of and formalize it in Coq, enabling us to prove
-lemmas and theorems about it.
+lemmas and theorems about it. See Figure~\ref{?} for the Coq implementation of the
+syntax and semantics.
+
+The Legion semantics are defined as a small step operational semantics, which
+result in non-deterministic orderings of memory reads and writes. By defining a
+notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
+can happen concurrently. This enables Treichler et al. to prove that if the runtime
+makes scheduling decisions based on region permissions in a particular way, serial 
+semantics are preserved with respect to the resulting heap state and return
+value. This is a desirable property: it means that any code using Legion can
+reason, either formally or informally, about its code with respect to a simple
+serial semantics, and have those semantics be preserved for parallel execution.
+This is in contrast to programming models like Cilk, which claim a weaker
+property: a serial execution is one of many valid executions.
 
@@
 behavior. Undefined behavior in a specification is well defined. A little bit
-like known unknowns vs. unknown unknowns, gaps in a specification
+like known unknowns vs. unknown unknowns, gaps in a specification can make it
+impossible to prove desired properties of a system.
 
@@
 
-Of course
+\subsection{Performance}
+
+Returning to the "P" in "HPC", an important area of research is formal
+reasoning about performance. One of the challenges for Legion is that the
+scheduler has a non-trivial amount of work to do. Generally at least one core
+per node is dedicated entirely to runtime overheads. This is one area where
+Cilk, though fairly informally, shines. They prove a number of bounds on work
+that the scheduler has to do.
+
+Ideally, we'd like to make similar claims about Legion runtime overheads.
+Currently, due to the complexity of the runtime implementation, this is
+infeasible. Like some of the desired correctness properties, a significant
+simplification of the runtime, and perhaps the semantics, would likely be
+required to achieve such a goal.
+
+\section{Conclusion}
+
+We hope to have convinced the reader of the importance and viability of proving
+software systems in HPC correct. While we've only taken small steps on the path
+to provably correct Legion, our hope is that it will inspire support for such 
+ventures by providing a convincing story on how to move forward. There is ever-
+increasing evidence that this class of deep specifications and certified software
+is an incredibly powerful tool for proving desirable properties about our large 
+software systems. We'd like to see more focus on correctness in HPC, and we think
+deep specifications and certified proofs are the way forward.
+
+ \bibliographystyle{ACM-Reference-Format}
+ \bibliography{annotated}
+\end{document}

### 2017-08-15 correctnesshpc/paper.tex
More writing changes

@@
 values it computes are correct? Presumably, these computations inform important
-decisions about topics ranging from climate to nuclear safety. We should be as
-confident as possible that we aren't basing these decisions on unexpected
-semantics.
+decisions in areas ranging from climate to nuclear safety. We should strive to
+be as confident as possible that we aren't basing these decisions on buggy
+code.
 
-Recent work has shown that modular, composable formal specifications enable
-formal correctness proofs for large software systems that were previously
-unattainable. By integrating a formal specification tightly with an
-implementation, one can achieve the highest levels of confidence in code
-currently possible \cite{?}. For example, the Compcert project implements an
-optimizing c compiler and proves it correct. The success of Compcert in
-removing compiler bugs has been well documented. We will discuss later how 
-Compcert could be used in building verifiable HPC systems.
+The strongest guarantees of correctness come from the area of formal specifications 
+and proofs. A specification is a description of how a program should behave, while proofs
+are used to show that an implementation abides by a specification.
+\emph{Formal} specifications and proofs are defined as specifications and
+proofs that have been embedded in a machine-checked formal logic \cite{Coq,
+isabelle...}. This process of formalizing specifications and proofs in machine
+checked logics has been shown to catch common errors in previously informal proofs.
+
+In addition, recent work has shown that modular, composable formal
+specifications enable formal correctness proofs for large software systems that
+were previously unattainable \cite{?}. By integrating a formal specification
+tightly with an implementation, one can achieve the highest levels of
+confidence in code currently possible \cite{?}. For example, the Compcert
+project implements an optimizing c compiler and proves it correct. The success
+of Compcert in removing compiler bugs has been well documented. We will discuss
+later how Compcert could be used in building verifiable HPC systems.
 
 In HPC, programming models have become an increasingly important way to ensure
-that code stays maintanable, performant, and portable. We argue that these
-programming models provide a good target for formalization: any correctness
-arguments can be leveraged by every program using the model. It is exactly this 
-kind of modularization of proofs that has been so successful in proving programs 
-correct \cite{?}.
+that code stays maintanable, performant, and portable \cite{?}. We claim that
+these programming models provide a good target for formalization: any
+correctness arguments can be leveraged by every program using the model. It is
+exactly this kind of modularization of proofs that has been so successful in
+proving programs correct \cite{?}.
 
@@
 Legion is a programming model for high performance computing that enables
-reasoning about locality and with the attractive property that it is guaranteed
-to preserve serial semantics. It achieves this by dividing memory into logical
-regions.  This allows programmers to avoid the
-burden
+reasoning about locality and concurrency via logical regions. It has the
+attractive property that it is guaranteed to preserve serial semantics for
+memory operations within logical regions. This allows programmers to avoid the
+burden of reasoning about the non-determinism that arises from concurrent
+computation. For this paper, we'll be basing our formalization work on the
+OOPSLA 2013 by Triechler et. al.  While there has been much work extending
+Legion since then, the core semantics has stayed relatively stable.
 
@@
 Specifications come in a wide range of formats. Many, like the OpenMP
-specification, are very informal, primarily consisting of english. Others, like
-Legion's previously, are written by hand in the using standard notation for
-logical relations, etc. For this paper, when we refer to a \emph{formal}
-specification, we are referring to a specification defined in a formal logic,
-like Coq.
+specification, are very informal, primarily consisting of english descriptions
+of how different constructs should behave. Others, like Legion's, are written
+by hand in the using standard notation for logical relations, etc. For this
+paper, when we refer to a \emph{formal} specification, we are referring to a
+specification defined in a formal logic, like Coq.
 
-While semantics defined in papers are a huge improvement over natural language
-specifications, there is still room for improvement. \cite{formalspec vs paper} has 
-shown that even hand-written mathematical specifications and proofs in computer
-science papers almost always have errors that can be caught by a formal proof.
-This is not to say, of course, that formal specifications and proofs are
-infallible.  On the contrary, it is quite easy to get a specification wrong.
-That said, often, if a there is a mistake in a spec , the process of proving it
-formally will expose that mistake, allowing for its correction.
+While semantics defined by hand as logical relations are a huge improvement
+over natural language specifications, there is still much room for improvement.
+\cite{formalspec vs paper} has shown that even hand-written mathematical
+specifications and proofs in computer science papers almost always have errors
+that can be caught by a formal proof.  This is not to say, of course, that
+formal specifications and proofs are infallible.  On the contrary, it is quite
+easy to get a specification wrong.  That said, often, if a there is a mistake
+in a spec , the process of proving it formally will expose that mistake,
+allowing for its correction.
 
@@
 "A program without a specification cannot be incorrect, it can only be
-surprising." \cite{appel}. One of the most important and challenging aspects of
-proving correctness is defining a formal specification of what a program
-*should* do. Indeed, there is currently a large NSF effort dedicated to
-improving our ability to do just this \cite{deepspec}. The choice of
-specification has far-reaching consequences in how easy it is to later prove
-correctness properties. For example, if a specification is chosen poorly,
+surprising." \cite{appel} One of the most important and challenging aspects of
+proving correctness is defining a formal specification of how a program
+should behave. Indeed, there is currently a large NSF effort involving many of
+the best minds in programming languages dedicated to improving our ability to
+do just this \cite{deepspec}. The choice of specification has far-reaching
+consequences in how easy it is to later prove correctness properties.
 
@@
 notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
-can happen concurrently. This enables Treichler et al. to prove that if the runtime
-makes scheduling decisions based on region permissions in a particular way, serial 
-semantics are preserved with respect to the resulting heap state and return
-value. This is a desirable property: it means that any code using Legion can
-reason, either formally or informally, about its code with respect to a simple
-serial semantics, and have those semantics be preserved for parallel execution.
-This is in contrast to programming models like Cilk, which claim a weaker
-property: a serial execution is one of many valid executions.
+can happen concurrently. This should allow us to prove formally that if the
+runtime makes scheduling decisions based on region permissions in a particular
+way, serial semantics are preserved with respect to the resulting heap state
+and return value. This is a desirable property: it means that any code using
+Legion can reason, either formally or informally, about its code with respect
+to a simple serial semantics, and have those semantics be preserved for
+parallel execution. This is in contrast to programming models like Cilk, which
+claim a weaker property: a serial execution is one of many valid executions.
 

### 2017-08-17 correctnesshpc/paper.tex
Added code to paper, reverted to default sigconf

@@
-\documentclass[sigplan]{acmart}
+\documentclass[sigconf]{acmart}
 
 \usepackage{booktabs} % For formal tables
-\usepackage{listings} % For formal tables
+
+% Dejavu fonts have unicode character support
+\usepackage{DejaVuSansMono}
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+\usepackage{listings} 
+\usepackage{fancyvrb}
+
+\usepackage{amssymb}
+\usepackage{bbm}
+\usepackage[greek,english]{babel}
+\usepackage{textgreek}
+
+%\usepackage{fontspec}
+%\defaultfontfeatures{Ligatures=TeX}
+
+\usepackage{ucs}
+\usepackage[utf8x]{inputenc}
+\usepackage{autofe}
 
@@
 %\setcopyright{acmlicensed}
-\settopmatter{printacmref=false}
+%\settopmatter{printacmref=false}
 \setcopyright{rightsretained}
@@
 
+\author{Pat McCormick}
+\affiliation{\institution{Los Alamos National Laboratories}}
+\email{pat@lanl.gov}
+
 \begin{abstract}
@@
 
-The strongest guarantees of correctness come from the area of formal specifications 
-and proofs. A specification is a description of how a program should behave, while proofs
-are used to show that an implementation abides by a specification.
-\emph{Formal} specifications and proofs are defined as specifications and
-proofs that have been embedded in a machine-checked formal logic \cite{Coq,
-isabelle...}. This process of formalizing specifications and proofs in machine
-checked logics has been shown to catch common errors in previously informal proofs.
+The strongest guarantees of correctness come from the area of formal
+specifications and proofs. A specification is a description of how a program
+should behave, while proofs are used to show that an implementation conforms to
+a specification. \emph{Formal} specifications and proofs are defined as
+specifications and proofs that have been embedded in a machine-checked formal
+logic \cite{Coq, isabelle...}. This process of formalizing specifications and
+proofs in machine checked logics has been shown to catch common errors in
+previously informal proofs \cite{papers}, and therefore is now considered
+to provide the highest assurance of correctness currently possible \cite{?}.
 
@@
 is no small task, as we discuss later, and the ways forward are numerous and
-lengthy. But if we believe that HPC informs important decisions, we must strive
-for this pinnacle of software correctness!
+difficult. But if we believe that HPC informs important decisions, we must
+rise to the challenge.
 
@@
 burden of reasoning about the non-determinism that arises from concurrent
-computation. For this paper, we'll be basing our formalization work on the
-OOPSLA 2013 by Triechler et. al.  While there has been much work extending
-Legion since then, the core semantics has stayed relatively stable.
+computation. For this paper, we'll be formalizating work on from Triecher et.
+al \cite{ooplsa13} While there has been much work extending Legion since then,
+the core semantics has stayed relatively stable \cite{regent, deppart}.
 
@@
 specification, are very informal, primarily consisting of english descriptions
-of how different constructs should behave. Others, like Legion's, are written
-by hand in the using standard notation for logical relations, etc. For this
-paper, when we refer to a \emph{formal} specification, we are referring to a
-specification defined in a formal logic, like Coq.
+of how different constructs should behave \cite{openmp}. Others, like Legion's,
+are written by hand in the using standard notation for logical relations, etc.
+For this paper, when we refer to a \emph{formal} specification, we are
+referring to a specification defined in a formal logic, like Coq.
 
@@
 
-"A program without a specification cannot be incorrect, it can only be
-surprising." \cite{appel} One of the most important and challenging aspects of
-proving correctness is defining a formal specification of how a program
-should behave. Indeed, there is currently a large NSF effort involving many of
-the best minds in programming languages dedicated to improving our ability to
-do just this \cite{deepspec}. The choice of specification has far-reaching
-consequences in how easy it is to later prove correctness properties.
+A program without a specification cannot be incorrect, it can only be
+surprising \footnote{Paraphrase of J.J. Horning, 1982}. One of the most
+important and challenging aspects of proving correctness is defining a formal
+specification of how a program should behave. Indeed, there is currently a
+large NSF effort involving many of the best minds in programming languages
+dedicated to improving our ability to do just this \cite{deepspec}. The choice
+of specification has far-reaching consequences in how easy it is to later prove
+correctness properties.
 
@@
 We take this specification of and formalize it in Coq, enabling us to prove
-lemmas and theorems about it. See Figure~\ref{?} for the Coq implementation of the
-syntax and semantics.
+lemmas and theorems about it. See Figure~\ref{?} for the Coq implementation of
+the syntax, and Figure~\ref{semantics}and semantics. All code is available
+online at \texttt{https://github.com/lanl/formallegion}.
+
+\begin{figure}
+\centering
+\begin{BVerbatim}
+Inductive e :=
+  | bv : bool → e
+  | iv : nat → e
+  | etuple : list e → e
+  | eindex : e → nat → e
+  | id : var → e
+  | new : T → r → e
+  | null : T → r → e
+  | isnull : e → e
+  | upr : e → list r → e
+  | downr : e → list r → e
+  | read : e → e 
+  | write : e → e → e
+  | reduce : var → e → e
+  | newcolor : r → e
+  | color : e → e → e → e
+  | add : e → e → e
+  | lt : e → e → e
+  | elet : var → T → e → e → e
+  | eif : e → e → e → e
+  | call : var → list r → list e → e
+  | partition : r → e → list r → e → e
+  | pack : e → T → list r → e
+  | unpack : e → var → T → list r → e → e.
+\end{BVerbatim}
+\caption{Legion expression syntax}
+\label{syntax}
+\end{figure}
 
@@
 
-\section{Discussion}
+\begin{figure*}
+\centering
+\begin{BVerbatim}
+Reserved Notation " input ↦ output " (at level 50).
+Inductive eval: Map r ρ 
+              * Map var v
+              * Map l T
+              * Map l v
+              * list l 
+              * e 
+              → v * list memop
+              → Prop := 
+  | ERead1 : ∀ M L H S C e l E v,
+    (M, L, H, S, C, e) ↦ (vl l, E) →
+    l ∉ C → 
+    lookup l (apply E S) = Some v → 
+    (M, L, H, S, C, read e) ↦ (v, E ++ [mread l v])
+  | ERead2 : ∀ M L H S C e l E v,
+    (M, L, H, S, C, e) ↦ (vl l, E) →
+    l ∈ C → 
+    (M, L, H, S, C, read e) ↦ (v, E ++ [mread l v]) 
+  | EWrite : ∀ M L H S S' C e1 e2 l E1 E2 E v,
+    (M, L, H, S, C, e1) ↦ (vl l, E1) →
+    (M, L, H, S', C, e2) ↦ (v, E2) →
+    valid_interleave S C E [E1; E2] →
+    (M, L, H, S, C, write e1 e2) ↦ (vl l, E ++ [mwrite l v])
+  | ENew : ∀ M L H S C t l r, 
+    l ∈ lu r M →
+    l ∉ domain S →
+    (M, L, H, S, C, new t r) ↦ (vl l, [])
+  | EUpRgn : ∀ M L H S C e v E rs,
+    (M, L, H, S, C, e) ↦ (v, E) →
+    (M, L, H, S, C, upr e rs) ↦ (v, E)
+  | EColor : ∀ M L H S C K E1 E2 E3 E l e1 e2 e3 v, 
+    (M, L, H, S, C, e1) ↦ (vcoloring K, E1) → 
+    (M, L, H, apply E1 S, C, e2) ↦ (vl l, E2) → 
+    (M, L, H, apply E2 (apply E1 S), C, e3) ↦ (viv v, E3) → 
+    valid_interleave S C E [E1; E2; E3] →
+    (M, L, H, S, C, color e1 e2 e3) ↦ (vcoloring ((l,v)::K), E) 
+  | EPartition : ∀ M M' L H S C rs rp ρs e1 e2 K E' E1 E2 v, 
+    (M, L, H, S, C, e1) ↦ (vcoloring K, E1) → 
+    ρs = map (map fst) (groupBy (λ x y, beq_nat (snd x) (snd y)) K) →
+    M' = zip rs ρs ++ M →  
+    (M', L, H, S, C, e2) ↦ (v, E2) →
+    valid_interleave S C E' [E1; E2] →
+    (M, L, H, S, C, partition rp e1 rs e2) ↦ (v, E')  
+  | ECall : ∀ M L H S C es vs Es xs id rs En1 E'' E' M' L' S' C' t ts rs' e Phi Q v, 
+    (∀ e v E, In (e, (v, E)) (zip es (zip vs Es)) → (M, L, H, S, C, e) ↦ (v, E)) →
+    valid_interleave S C E' Es →
+    M' = zip rs' (map (λ r, lu r M) rs) →  
+    L' = zip xs vs →
+    S' = apply E' S → 
+    function id rs xs ts Phi Q t e →
+    (M', L', H, S', C', e) ↦ (v, En1) → 
+    valid_interleave S C E'' [E'; En1] →
+    (M, L, H, S, C, call id rs es) ↦ (v, E'')
+    ...
+where " input ↦ output " := (eval input output).
+\end{BVerbatim}
+\caption{Fragment of Legion's formal semantics defined in Coq}
+\label{semantics}
+\end{figure*}
 
-One common theme in formalizing systems is the discovery of holes in the
+The language is made up of some standard elements, e.g. \texttt{call},
+\texttt{id}, and some Legion-specific constructs e.g. \texttt{partition},
+\texttt{upr}. Evaluation is similar in nature, with the standard elements
+having standard semantics, and the
+
+\section{Proofs}
+
+While so far we have discussed the formalization of the \emph{specification},
+we have largely left discussion of proofs by the wayside. In this section, we
+discuss the challenges of proving useful properties of real world systems, and
+give some examples of basic lemmas that will be useful in proving correctness
+properties of Legion. The hope is that the reader will gain some insight into
+the challenges involved in proving formal properties of systems of real-world
+systems.
+
+One of the core notions of Legion is the notion of region relations. These
+describe how regions are related, either by being disjoint, or by being
+subregions of one another. See Figure~\ref{regionrel} for the syntax.
+
+\begin{figure*}
+\centering
+\begin{BVerbatim}
+Inductive ω := 
+  | subr : r → r → ω 
+  | disj : r → r → ω.
+Definition Ω := set ω.
+
+Reserved Notation " ω ∈* Ω " (at level 40).
+Inductive cclo (Ω : Ω) : ω → Prop := 
+  | trivial : ∀ ω, ω ∈ Ω → ω ∈* Ω
+  | sub_refl : ∀ r, subr r r ∈* Ω
+  | sub_trans : ∀ ri rj rk, subr ri rj ∈* Ω → subr rj rk ∈* Ω  → 
+                                  subr ri rk ∈* Ω 
+  | dis_sub : ∀ ri rj rk, subr ri rj ∈* Ω → disj rj rk ∈*  Ω → 
+                                  disj ri rk ∈* Ω
+  | dis_comm : ∀ ri rj, disj ri rj ∈* Ω → disj rj ri ∈* Ω 
+where " ω ∈* Ω " := (cclo Ω ω).
+
+Definition resp (M : M) (elem : Ω → ω → Prop) (Ω : Ω) :=
+  (∀ r1 r2, elem Ω (subr r1 r2) → lu r1 M ⊆ lu r2 M)
+                              ∧ 
+  (∀ r1 r2, elem Ω (disj r1 r2) → lu r1 M ∩ lu r2 M = [])
+.
+
+Notation " M ~ Ω " := (resp M (λ Ω ω, In ω Ω) Ω) (at level 40).
+Notation " M ~* Ω " := (resp M cclo Ω) (at level 40).
+\end{BVerbatim}
+\caption{Syntax for region relations}
+\label{regionrel}
+\end{figure*}
+
+One important property when proving correctness in Legion is to show that the
+relations are equivalent to their closure, with respect to what they imply
+about locations in those regions.  Figure~\ref{regionrelproof} for the
+statement of this proof. Intuitively, this means that if a region is a
+subregion of another region, and that region is a subregion of a third, then
+the first subregion is a subregion of the third. Disjointness is commutative, 
+and subregions of disjoint regions are disjoint.
+
+\begin{figure}
+\centering
+\begin{BVerbatim}
+Lemma resp_clos_sub : ∀ r1 r2 Ω (M:M), 
+  (∀ r1 r2, subr r1 r2 ∈ Ω → lu r1 M ⊆ lu r2 M) → 
+           subr r1 r2 ∈* Ω → lu r1 M ⊆ lu r2 M.
+...
+
+Lemma resp_clos : ∀ M Ω, M ~ Ω ↔ M ~* Ω.
+...
+\end{BVerbatim}
+\caption{Lemmas about region relations}
+\label{regionrelproof}
+\end{figure}
+
+TODO: Noah, if you want to put a rough description of your preservation proof 
+here feel free. Your type inference isn't loading for me.
+
+\section{Future Work}
+
+The work presented in this paper is a small fraction of a small fraction of the
+amount of work required to build a complete, correct-by-construction HPC
+application. In this section, we outline some of the possible paths towards
+this worthy goal.
+
+As mentioned earlier in the paper, one of the keystones for large formally
+correct systems is modularity. This modularity has two purposes. On the one
+hand, it enables ease of implementation. This has long been recognized 
+in the field of software engineering \cite{?}. The second, analagous advantage,
+is that in a similar way modularity eases the burden of proofs. By defining 
+specifications as abstractions, one can separate concerns of proofs. For example,
+by composing a proof that a program's serial semantics comply with a formal
+specification with a proof that a parallel language respects those serial 
+semantics, we can get a proof that a parallel program respects the original
+formal semantics. In HPC, by applying this idea to parallel programming models like
+Legion, we can separate the application developer from having to prove 
+correctness with respect to one or more increasingly complex, heterogeneous
+architechures. Instead, we can provide an intermediate layer, like Legion, that
+allows application developers to reason and prove correctness to much 
+simpler operational semantics, and then prove that those semantics are retained
+even on the complex, heterogeneous architectures of todays supercomputers.
+
+Of course this is much easier said than done, and we address some of the challenges
+faced by such an effort in the following subsections.
+
+\subsection{Current Limitations}
+
+One common theme in formalizing systems is the discovery of issues in the
 corresponding informal specification. For example, when formalizing the C
 language for the CompCert project, Leroy et. al. found many cases where the
-specification was incomplete. Note that this is seperate notion from undefined
-behavior. Undefined behavior in a specification is well defined. A little bit
-like known unknowns vs. unknown unknowns, gaps in a specification can make it
-impossible to prove desired properties of a system.
+informal specification was simply incomplete \cite{compcert}. Note that this is
+seperate notion from undefined behavior. Like unknown unknowns, gaps in a
+specification can make it impossible to prove desired properties of a system.
+In contrast, undefined behaviour is only a problem when programs exihibit it.
 
 In formalizing Legion, we discovered that while there is an impressive theory
-ensuring that the side effects of function calls running concurrently preserve
-serial semantics, there is no theory for what function calls can run
-concurrently based on value dependencies. For example, consider the infamous
-Fibonacci example, which naively computes the \texttt{n}'th Fibonacci number:
+ensuring that the side effects of concurrent evaluations preserve serial
+semantics with respect to memory reads and writes, there is no theory for what
+expressions can be evaluated concurrently based on value dependencies. For
+example, if we define the program
+\begin{center}
+\begin{BVerbatim}
+Definition read_write x y t T r :=
+  elet x t (read (id y)) 
+    (write (new T r) (id x)).
+\end{BVerbatim}
+\end{center}
+then the semantics will happily order the \texttt{read} \emph{after} the
+\texttt{write}, despite the fact that the value being written, \texttt{x},
+generally will need to be read before being written. This isn't catastrophic
+for the semantics, as the returned value and resulting heap are equivalent, 
+as desired, but it limits the ability to reason effectively about concurrency. 
+In the actual implementations this is handled both dynamically with futures
+and statically with dependency analysis \cite{regent}.
 
-\begin{lstlisting}
-fib(n) = if n < 2 then 1 else fib(n-1) + fib(n-2)
-\end{lstlisting}
+Another issue with the semantics as they are now is that there is no way to 
+constrain ordering on side effects \emph{other than reads and writes into
+regions}. For example, generally programs will need to make system calls or
+foreign function calls in a particular order, and currently in the semantics
+there is no support for enforcing such an ordering.
+
+A third issue is that there is no way to dynamically create and destroy
+regions. In other words, it the semantics assumes static memory allocation.
+Of course, this differs from the actual implementation, and is an important,
+though difficult, semantics to formalize. As memory heirarchies get more 
+sophisticated, the need for tools for reasoning about memory use and locality
+becomes more essential.
+
+A fourth challenge is that in practice, Legion is not a stand-alone language.
+Instead, it is embedded in an existing language. For the current implementations, 
+those languages are Regent and C/C++ \cite{regent, legion}. There are two 
+interesting paths towards addressing this issue. One would be to build a sort 
+of abstraction for languages that Legion is embedded in, and prove correctness
+with respect to that abstraction. This would likely be difficult to get right, as 
+capturing the important semantics of the languages that it would be embedded in
+would likely be a significant challenge. A second possibility for addressing this 
+issue would be choosing a particular semantics and focusing on that. In particular,
+combining the C API with the CompCert verified C work would likely be a
+fruitful, though challenging, venture.
+
+All of these issues are addressable given enough time and effort. Some offer
+a nice opportunity for modularity within the tasking formalization. For example, 
+ordering tasks based on dependencies of values is common accross all tasking
+systems \cite{cilk, hpx}, and is likely a sort of sub-semantics that they could
+all share, reducing proof burden across different runtimes.
 
@@
 
-Returning to the "P" in "HPC", an important area of research is formal
-reasoning about performance. One of the challenges for Legion is that the
-scheduler has a non-trivial amount of work to do. Generally at least one core
-per node is dedicated entirely to runtime overheads. This is one area where
-Cilk, though fairly informally, shines. They prove a number of bounds on work
-that the scheduler has to do.
+Returning to the "P" in "HPC", one important current area of research is formal
+reasoning about performance \cite{performance}. One of the challenges for
+Legion, and runtime systems in general, is that the scheduler often has a
+non-trivial amount of work to do, leading to large overheads. In Legion, the
+combination of reasoning about permissions, memory locality, etc., makes this
+overhead particularly hard to predict.
+
+This is one oft-overlooked advantage to formal reasoning: it allows reasoning
+about \emph{any} property of a system, not just correctness. For example, there
+is the work by \cite{?} working towards formally verified complexity of code.
+Chlipala et. al. have also used techniques like those described here to construct
+and verify hardware description, where performance must be reasoned about down
+to single cycles \cite{chlipala}.  This is one area where other tasking
+systems, like Cilk, have a better story than Legion. Blumofe et al. informally
+describe bounds on runtime overheads, which has been realized in practice as
+low overheads in the actual runtime \cite{cilk}.
 
 Ideally, we'd like to make similar claims about Legion runtime overheads.
-Currently, due to the complexity of the runtime implementation, this is
-infeasible. Like some of the desired correctness properties, a significant
-simplification of the runtime, and perhaps the semantics, would likely be
-required to achieve such a goal.
+Given a formal model of the work the runtime system has to do, one can reason
+about bounds on overhead. Currently, due to the complexity of the runtime
+implementation, this is infeasible. Like some of the desired correctness
+properties, a significant simplification of the runtime, and perhaps the
+semantics, would likely be required to achieve such a goal.
+
+\subsection{Other Tasking Systems}
+
+While we've focused on Legion, there are other parallel programming systems
+that offer exciting opportunities for formalization. While some (looking at
+you, OpenMP), are likely much to large and complex to formalize without
+herculean effort, others have simple enough semantics that formalization is
+likely within reach. One in particular that is attractive due to its simplicity
+is the recent work on parallel extensions to LLVM, the Tapir project
+\cite{tapir}. Building on existing formalization efforts for LLVM
+\cite{verillvm}, extending existing semantics with the Tapir instructions for
+concurrency should be viable. Again, there would be many challenges,
+particularly in the area of formal reasoning about memory models, but the
+approach should be viable.
+
+As mentioned earlier, there is also the important notion of \emph{shared
+semantics accross different tasking systems}. This notion of shared semantics
+presents a valuable opportunity for modularization of specification. By defining
+a shared semantics developers could reason more effectively, both formally and
+informally, about the semantics of their code with respect to tasking layers.
 
@@
 We hope to have convinced the reader of the importance and viability of proving
-software systems in HPC correct. While we've only taken small steps on the path
-to provably correct Legion, our hope is that it will inspire support for such 
-ventures by providing a convincing story on how to move forward. There is ever-
-increasing evidence that this class of deep specifications and certified software
-is an incredibly powerful tool for proving desirable properties about our large 
-software systems. We'd like to see more focus on correctness in HPC, and we think
-deep specifications and certified proofs are the way forward.
+parallel programming systems in HPC correct. While we've only taken small steps
+on the path to provably correct Legion, our hope is that it will inspire
+support for such ventures by providing a convincing story on how to move
+forward. There is ever- increasing evidence that this class of deep
+specifications and certified software is an incredibly powerful tool for
+proving desirable properties about our large software systems. We'd like to see
+more focus on correctness in HPC, and we think deep specifications and
+certified proofs are the way forward.
 
@@
 \bibliography{annotated}
+
 \end{document}

### 2017-08-17 correctnesshpc/paper.tex
Added a little more background

@@
 computation. For this paper, we'll be formalizating work on from Triecher et.
-al \cite{ooplsa13} While there has been much work extending Legion since then,
+al \cite{ooplsa13}. While there has been much work extending Legion since then,
 the core semantics has stayed relatively stable \cite{regent, deppart}.
 
+By partitioning regions into \emph{subregions}, and tracking \emph{privileges}, 
+Legion is able to ensure that certain code won't interfere or alias, and can
+therefor be safely, at least with respect to those memory regions, run in
+parallel. This theory has been put into practice to impressive effect, and has
+been shown to scale both within a node and between \cite{legionpapers}. That
+said, there are still challenges subtle bugs that occasionally show up in the
+implementation. It is possible that these kinds of bugs could manifest
+themselves in ways that would lead to incorrect conclusions when informing
+major decisions, something we'd generally like to avoid. One goal of this paper
+is to move towards formally and systematically removing \emph{all} bugs from
+the programming model, giving everyone who uses is far more certainty that 
+their code is correct.
+
+To move towards this goal, we turn to mechanized proof assistants.
 Improvements to proof assistants over recent decades have made them a powerful
@@
 
-Coq is a dependently typed proof assistant. At its core is the Calculus of
-Inductive Constructions, a small dependently typed programming language and
-logic, capable of powerful abstractions for formal proofs. In the last 10 years 
-great strides have been made in using it to both build systems and prove useful
-properties about them \cite{compcert, certicoq, chlipala...}.
+Coq, in particular, has proven itself to be a particularly potent dependently
+typed proof assistant. At its core is the Calculus of Inductive Constructions,
+a small dependently typed programming language and logic, capable of powerful
+abstractions for formal proofs \cite{cic}. In the last 10 years great strides
+have been made in using it to both build systems and prove useful properties
+about them \cite{compcert, certicoq, chlipala...}.
 
@@
 formal specifications and proofs are infallible.  On the contrary, it is quite
-easy to get a specification wrong.  That said, often, if a there is a mistake
-in a spec , the process of proving it formally will expose that mistake,
-allowing for its correction.
+easy to get a specification wrong. That said, often, if a there is a mistake in
+a spec , the process of proving it formally will expose that mistake, allowing
+for its correction.
 
@@
 We take this specification of and formalize it in Coq, enabling us to prove
-lemmas and theorems about it. See Figure~\ref{?} for the Coq implementation of
-the syntax, and Figure~\ref{semantics}and semantics. All code is available
-online at \texttt{https://github.com/lanl/formallegion}.
+lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
+implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
+is available online at \texttt{https://github.com/lanl/formallegion}.
 
@@
 
-The Legion semantics are defined as a small step operational semantics, which
+The Legion semantics are defined as a big step operational semantics, which
 result in non-deterministic orderings of memory reads and writes. By defining a
@@
 the first subregion is a subregion of the third. Disjointness is commutative, 
-and subregions of disjoint regions are disjoint.
+and subregions of disjoint regions are disjoint. We've left out the Coq tactic
+proofs for space, as they are likely less informative to the reader than the
+propositions and definitions, but they can be found online as linked above. 
+Lemma \texttt{resp\_clos\_sub} proves that given that if a subregion relation
+holds with respect to a mapping \texttt{M} and a set of region constraints
+\textOmega{Ω} for all regions \texttt{r1, r2} then holds for the closure of that
+set of region constraints under that mapping. Given \texttt{resp\_clos\_sub}, 
+Lemma \texttt{resp\_clos} proves that this implies that a mapping respects
+a set of region constraints if and onlyo if it respects the closure of that
+set. This non-trivial property is crucial for proving that the final decisions 
+made by the tasking system based on the sets of constraints preserves 
+serial semantics.
 

### 2017-08-24 correctnesshpc/paper.tex
Small changes

@@
 describe how regions are related, either by being disjoint, or by being
-subregions of one another. See Figure~\ref{regionrel} for the syntax.
+subregions of one another. See Figure~\ref{regionrel} for the syntax. Another
+important construction are mappers, \texttt{M}, which map logical regions
+\texttt{r} with a set of physical locations. We include some definitions as
+well, notably the notion that a mapping respects (\texttt{resp}) a set of
+region requirements iff for every subregion relation and disjoint relation, the
+mapping to sets of physical locations follows the corresponding set relation.
 
@@
 
-TODO: Noah, if you want to put a rough description of your preservation proof 
-here feel free. Your type inference isn't loading for me.
+TODO: Noah, if you want to put any other proof stuff here feel free.
 

### 2017-08-25 correctnesshpc/paper.tex
a few spelling/style fixes

@@
 confidence in code currently possible \cite{?}. For example, the Compcert
-project implements an optimizing c compiler and proves it correct. The success
+project implements an optimizing C compiler and proves it correct. The success
 of Compcert in removing compiler bugs has been well documented. We will discuss
@@
 In HPC, programming models have become an increasingly important way to ensure
-that code stays maintanable, performant, and portable \cite{?}. We claim that
+that code stays maintainable, performant, and portable \cite{?}. We claim that
 these programming models provide a good target for formalization: any
@@
 Legion is able to ensure that certain code won't interfere or alias, and can
-therefor be safely, at least with respect to those memory regions, run in
+therefore be safely, at least with respect to those memory regions, run in
 parallel. This theory has been put into practice to impressive effect, and has
 been shown to scale both within a node and between \cite{legionpapers}. That
-said, there are still challenges subtle bugs that occasionally show up in the
+said, there are still challenges --subtle bugs that occasionally show up in the
 implementation. It is possible that these kinds of bugs could manifest
@@
 For this paper, when we refer to a \emph{formal} specification, we are
-referring to a specification defined in a formal logic, like Coq.
+referring to a specification defined in a formal logic, eg. Coq.
 
@@
 systems \cite{cilk, hpx}, and is likely a sort of sub-semantics that they could
-all share, reducing proof burden across different runtimes.
+all share, reducing proof burden across different run-times.
 

### 2017-08-25 correctnesshpc/paper.tex
Small tweak to LANL org (lab vs. labs).
add citations

@@
 \author{George Stelle}
-\affiliation{\institution{Los Alamos National Laboratories}}
+\affiliation{\institution{Los Alamos National Laboratory}}
 \email{stelleg@lanl.gov}
@@
 \author{Pat McCormick}
-\affiliation{\institution{Los Alamos National Laboratories}}
+\affiliation{\institution{Los Alamos National Laboratory}}
 \email{pat@lanl.gov}
@@
 specifications and proofs that have been embedded in a machine-checked formal
-logic \cite{Coq, isabelle...}. This process of formalizing specifications and
+logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system, jackson1994nuprl}. This process of formalizing specifications and
 proofs in machine checked logics has been shown to catch common errors in
-previously informal proofs \cite{papers}, and therefore is now considered
+previously informal proofs \cite{clarke1996formal}, and therefore is now considered
 to provide the highest assurance of correctness currently possible \cite{?}.
@@
 specifications enable formal correctness proofs for large software systems that
-were previously unattainable \cite{?}. By integrating a formal specification
+were previously unattainable \cite{weng2016deepspec}. By integrating a formal specification
 tightly with an implementation, one can achieve the highest levels of
-confidence in code currently possible \cite{?}. For example, the Compcert
+confidence in code currently possible \cite{weng2016deepspec}. For example, the Compcert
 project implements an optimizing C compiler and proves it correct. The success
@@
 In HPC, programming models have become an increasingly important way to ensure
-that code stays maintainable, performant, and portable \cite{?}. We claim that
+that code stays maintainable, performant, and portable \cite{diaz2012survey}. We claim that
 these programming models provide a good target for formalization: any
@@
 exactly this kind of modularization of proofs that has been so successful in
-proving programs correct \cite{?}.
+proving programs correct \cite{weng2016deepspec}.
 
@@
 computation. For this paper, we'll be formalizating work on from Triecher et.
-al \cite{ooplsa13}. While there has been much work extending Legion since then,
-the core semantics has stayed relatively stable \cite{regent, deppart}.
+al \cite{treichler2013language}. While there has been much work extending Legion since then,
+the core semantics has stayed relatively stable \cite{slaughter2015regent, deppart}.
 
@@
 parallel. This theory has been put into practice to impressive effect, and has
-been shown to scale both within a node and between \cite{legionpapers}. That
+been shown to scale both within a node and between \cite{bauer2012legion, treichler2014realm}. That
 said, there are still challenges --subtle bugs that occasionally show up in the
@@
 a small dependently typed programming language and logic, capable of powerful
-abstractions for formal proofs \cite{cic}. In the last 10 years great strides
+abstractions for formal proofs \cite{bertot2013interactive}. In the last 10 years great strides
 have been made in using it to both build systems and prove useful properties
-about them \cite{compcert, certicoq, chlipala...}.
+about them \cite{leroy2012compcert, anand2017certicoq, chlipala2009effective}.
 
@@
 specification, are very informal, primarily consisting of english descriptions
-of how different constructs should behave \cite{openmp}. Others, like Legion's,
+of how different constructs should behave \cite{dagum1998openmp}. Others, like Legion's,
 are written by hand in the using standard notation for logical relations, etc.
@@
 large NSF effort involving many of the best minds in programming languages
-dedicated to improving our ability to do just this \cite{deepspec}. The choice
+dedicated to improving our ability to do just this \cite{weng2016deepspec}. The choice
 of specification has far-reaching consequences in how easy it is to later prove
@@
 One thing that makes Legion unique among high performance runtime systems is
-that it provides a specification of its operational semantics \cite{oopsla13}.
+that it provides a specification of its operational semantics \cite{treichler2013language}.
 We take this specification of and formalize it in Coq, enabling us to prove
@@
 ordering tasks based on dependencies of values is common accross all tasking
-systems \cite{cilk, hpx}, and is likely a sort of sub-semantics that they could
+systems \cite{blumofe1995cilk, kaiser2014hpx}, and is likely a sort of sub-semantics that they could
 all share, reducing proof burden across different run-times.
@@
 and verify hardware description, where performance must be reasoned about down
-to single cycles \cite{chlipala}.  This is one area where other tasking
+to single cycles \cite{braibant2013formal}.  This is one area where other tasking
 systems, like Cilk, have a better story than Legion. Blumofe et al. informally
 describe bounds on runtime overheads, which has been realized in practice as
-low overheads in the actual runtime \cite{cilk}.
+low overheads in the actual runtime \cite{blumofe1995cilk}.
 
@@
 is the recent work on parallel extensions to LLVM, the Tapir project
-\cite{tapir}. Building on existing formalization efforts for LLVM
-\cite{verillvm}, extending existing semantics with the Tapir instructions for
+\cite{schardl2017tapir}. Building on existing formalization efforts for LLVM
+\cite{zhao2012formalizing}, extending existing semantics with the Tapir instructions for
 concurrency should be viable. Again, there would be many challenges,

### 2017-08-26 correctnesshpc/paper.tex
draft of progress and preservation theorems

@@
 
-TODO: Noah, if you want to put any other proof stuff here feel free.
+\subsection{Full Implementation of the Specification}
+In addition to proving theorems and lemmas about the semantics of Legion as they
+relate to HPC memory models, another goal of a formal specification is to ensure
+both the completeness and soundness of the Core Legion model. Much of the
+necessary implementation work here is ensuring that the entire grammar of the
+language is both type safe and its operational semantics contain transition
+rules for all productions in the language.
+
+To do this we are writing our proofs of progress and preservation (ie. that well
+typed expressions evaluate to a value or diverge and that
+every transition in the operational semantics produces a well typed value given
+a well typed expression) together with our implementation of the semantics of
+the Core Legion language.
+
+Thus far this approach has acted as a sanity check for our implementation and helped us
+handle unspecified corner cases in the Core Legion specification (eg. the list
+of functions in the \texttt{T-Program} relation has no corresponding transition in the
+operational semantics).
+
+The theorems we are using are described in Figure~\ref{preservationtheorem} and
+Figure~\ref{progresstheorem} (see
+\texttt{https://github.com/lanl/formallegion} for the in progress proofs).
+
+\begin{figure}
+\begin{BVerbatim}
+Theorem preservation : forall (e5:e) (t:T) (E:E) (M:M) (L:L) 
+(H:H) (S:list (e * e)) (C:list e) (T:T),
+    exists v,
+      eval M L H S C e5 v E ->
+      typing EmptyContext EmptyPriv EmptyConstraint e5 T  ->
+      exists G P Omega T, typing G P Omega v T.
+\end{BVerbatim}
+\caption{preservation theorem}
+\label{preservationtheorem}
+\end{figure}
+
+\begin{figure}
+\begin{BVerbatim}
+Theorem progress : forall  (e5:e) (t:T) (M:M) (H:H)  
+(C:list e) (T1 T2:T) L (S:list (e * e)),
+    typing EmptyContext EmptyPriv EmptyConstraint e5 T1  ->
+    (exists E v ,
+        is_value v ->
+        eval M L H S C e5 v E).
+\end{BVerbatim}
+\caption{progress theorem}
+\label{progresstheorem}
+\end{figure}
 

### 2017-08-26 correctnesshpc/paper.tex
Editing pass. Citations complete.

@@
 specifications and proofs that have been embedded in a machine-checked formal
-logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system, jackson1994nuprl}. This process of formalizing specifications and
-proofs in machine checked logics has been shown to catch common errors in
-previously informal proofs \cite{clarke1996formal}, and therefore is now considered
-to provide the highest assurance of correctness currently possible \cite{?}.
+logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
+jackson1994nuprl}. This process of formalizing specifications and proofs in
+machine checked logics has been shown to catch common errors in previously
+informal proofs, and therefore is often considered to provide the highest
+assurance of correctness currently possible \cite{clarke1996formal,
+yang2011finding}.
 
@@
 specifications enable formal correctness proofs for large software systems that
-were previously unattainable \cite{weng2016deepspec}. By integrating a formal specification
-tightly with an implementation, one can achieve the highest levels of
-confidence in code currently possible \cite{weng2016deepspec}. For example, the Compcert
-project implements an optimizing C compiler and proves it correct. The success
-of Compcert in removing compiler bugs has been well documented. We will discuss
-later how Compcert could be used in building verifiable HPC systems.
+were previously unattainable \cite{weng2016deepspec}. By integrating a formal
+specification tightly with an implementation, one can achieve the highest
+levels of confidence in code currently possible \cite{weng2016deepspec}. For
+example, the Compcert project implements an optimizing C compiler and proves it
+correct. The success of Compcert in removing compiler bugs has been well
+documented. We will discuss later how Compcert could be used in building
+verifiable HPC systems.
 
 In HPC, programming models have become an increasingly important way to ensure
-that code stays maintainable, performant, and portable \cite{diaz2012survey}. We claim that
-these programming models provide a good target for formalization: any
-correctness arguments can be leveraged by every program using the model. It is
-exactly this kind of modularization of proofs that has been so successful in
+that code stays maintainable, performant, and portable \cite{diaz2012survey}.
+We claim that these programming models provide a good target for formalization:
+any correctness arguments can be leveraged by every program using the model. It
+is exactly this kind of modularization of proofs that has been so successful in
 proving programs correct \cite{weng2016deepspec}.
@@
 the kinds of things one can prove in an HPC settings. Finally, we discuss the
-some of the possible paths towards full, formally correct implementation. This
-is no small task, as we discuss later, and the ways forward are numerous and
-difficult. But if we believe that HPC informs important decisions, we must
+some of the possible paths towards a full, formally correct implementation.
+This is no small task, as we discuss later, and the ways forward are numerous
+and difficult. But if we believe that HPC informs important decisions, we must
 rise to the challenge.
@@
 Legion is a programming model for high performance computing that enables
-reasoning about locality and concurrency via logical regions. It has the
-attractive property that it is guaranteed to preserve serial semantics for
+reasoning about locality and concurrency via logical regions of memory. It has
+the attractive property that it is guaranteed to preserve serial semantics for
 memory operations within logical regions. This allows programmers to avoid the
@@
 computation. For this paper, we'll be formalizating work on from Triecher et.
-al \cite{treichler2013language}. While there has been much work extending Legion since then,
-the core semantics has stayed relatively stable \cite{slaughter2015regent, deppart}.
+al \cite{treichler2013language}. While there has been much work extending
+Legion since then, the core semantics has stayed relatively stable
+\cite{slaughter2015regent, treichler2016dependent}.
 
@@
 parallel. This theory has been put into practice to impressive effect, and has
-been shown to scale both within a node and between \cite{bauer2012legion, treichler2014realm}. That
-said, there are still challenges --subtle bugs that occasionally show up in the
-implementation. It is possible that these kinds of bugs could manifest
-themselves in ways that would lead to incorrect conclusions when informing
-major decisions, something we'd generally like to avoid. One goal of this paper
-is to move towards formally and systematically removing \emph{all} bugs from
-the programming model, giving everyone who uses is far more certainty that 
-their code is correct.
+been shown to scale both within a node and between \cite{bauer2012legion,
+treichler2014realm}. That said, there are still subtle bugs that occasionally
+show up in the implementation. One goal of this paper is to move towards
+formally and systematically removing \emph{all} bugs from the programming
+model, giving everyone who uses is far more certainty that their code is
+correct.
 
@@
 a small dependently typed programming language and logic, capable of powerful
-abstractions for formal proofs \cite{bertot2013interactive}. In the last 10 years great strides
-have been made in using it to both build systems and prove useful properties
-about them \cite{leroy2012compcert, anand2017certicoq, chlipala2009effective}.
+abstractions for formal proofs \cite{bertot2013interactive}. In the last 10
+years great strides have been made in using it to both build systems and prove
+useful properties about them \cite{leroy2012compcert, anand2017certicoq,
+chlipala2009effective}.
 
@@
 specification, are very informal, primarily consisting of english descriptions
-of how different constructs should behave \cite{dagum1998openmp}. Others, like Legion's,
-are written by hand in the using standard notation for logical relations, etc.
-For this paper, when we refer to a \emph{formal} specification, we are
-referring to a specification defined in a formal logic, eg. Coq.
+of how different constructs should behave \cite{dagum1998openmp}. Others, like
+Legion's, are written by hand in the using standard notation for logical
+relations. For this paper, when we refer to a \emph{formal} specification, we
+are referring to a specification defined in a formal logic, eg. Coq.
 
@@
 over natural language specifications, there is still much room for improvement.
-\cite{formalspec vs paper} has shown that even hand-written mathematical
-specifications and proofs in computer science papers almost always have errors
-that can be caught by a formal proof.  This is not to say, of course, that
-formal specifications and proofs are infallible.  On the contrary, it is quite
-easy to get a specification wrong. That said, often, if a there is a mistake in
-a spec , the process of proving it formally will expose that mistake, allowing
-for its correction.
+Like type checking a program, type checking a proof exposes many "bugs" in a
+proof. This is not to say, of course, that formal specifications and proofs are
+infallible. On the contrary, it is quite easy to get a specification wrong.
+That said, often, if a there is a mistake in a spec, the process of proving it
+formally will expose that mistake, allowing for its correction.
 
@@
 lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
-implementation of the syntax, and Figure~\ref{semantics}and semantics. All code
-is available online at \texttt{https://github.com/lanl/formallegion}.
+implementation of the syntax, and Figure~\ref{semantics} for semantics. All
+code is available online at \texttt{https://github.com/lanl/formallegion}.
 
@@
 \texttt{upr}. Evaluation is similar in nature, with the standard elements
-having standard semantics, and the
+having standard semantics with the addition that memory reads and writes must
+be valid interleaves of subexpression memory operations. In simpler terms, only
+orderings of memory operations that preserve the resulting memory structure are
+permitted.
 
@@
            subr r1 r2 ∈* Ω → lu r1 M ⊆ lu r2 M.
-...
 
 Lemma resp_clos : ∀ M Ω, M ~ Ω ↔ M ~* Ω.
-...
 \end{BVerbatim}
@@
 
-\subsection{Full Implementation of the Specification}
-In addition to proving theorems and lemmas about the semantics of Legion as they
-relate to HPC memory models, another goal of a formal specification is to ensure
-both the completeness and soundness of the Core Legion model. Much of the
-necessary implementation work here is ensuring that the entire grammar of the
-language is both type safe and its operational semantics contain transition
+\subsection{Specification Properties}
+In addition to proving theorems and lemmas about the semantics of Legion as
+they relate to HPC memory models, another goal of a formal specification is to
+ensure both the completeness and soundness of the Core Legion model. Much of
+the necessary implementation work here is ensuring that the entire grammar of
+the language is both type safe and its operational semantics contain transition
 rules for all productions in the language.
@@
 To do this we are writing our proofs of progress and preservation (ie. that well
-typed expressions evaluate to a value or diverge and that
-every transition in the operational semantics produces a well typed value given
-a well typed expression) together with our implementation of the semantics of
-the Core Legion language.
+typed expressions evaluate to a value or diverge and that every transition in
+the operational semantics produces a well typed value given a well typed
+expression) together with our implementation of the semantics of the Core
+Legion language.
 
-Thus far this approach has acted as a sanity check for our implementation and helped us
-handle unspecified corner cases in the Core Legion specification (eg. the list
-of functions in the \texttt{T-Program} relation has no corresponding transition in the
-operational semantics).
+Thus far this approach has acted as a sanity check for our implementation and
+helped us handle unspecified corner cases in the Core Legion specification (eg.
+the list of functions in the \texttt{T-Program} relation has no corresponding
+transition in the operational semantics).
 
 The theorems we are using are described in Figure~\ref{preservationtheorem} and
-Figure~\ref{progresstheorem} (see
-\texttt{https://github.com/lanl/formallegion} for the in progress proofs).
+Figure~\ref{progresstheorem} (see \texttt{https://github.com/lanl/formallegion}
+for the in progress proofs).
 
@@
 correct systems is modularity. This modularity has two purposes. On the one
-hand, it enables ease of implementation. This has long been recognized 
-in the field of software engineering \cite{?}. The second, analagous advantage,
-is that in a similar way modularity eases the burden of proofs. By defining 
-specifications as abstractions, one can separate concerns of proofs. For example,
-by composing a proof that a program's serial semantics comply with a formal
-specification with a proof that a parallel language respects those serial 
-semantics, we can get a proof that a parallel program respects the original
-formal semantics. In HPC, by applying this idea to parallel programming models like
-Legion, we can separate the application developer from having to prove 
-correctness with respect to one or more increasingly complex, heterogeneous
-architechures. Instead, we can provide an intermediate layer, like Legion, that
-allows application developers to reason and prove correctness to much 
-simpler operational semantics, and then prove that those semantics are retained
-even on the complex, heterogeneous architectures of todays supercomputers.
+hand, it enables ease of implementation. The second, analagous advantage, is
+that in a similar way modularity eases the burden of proofs. By defining
+specifications as abstractions, one can separate concerns of proofs. For
+example, by composing a proof that a program's serial semantics comply with a
+formal specification with a proof that a parallel language respects those
+serial semantics, we can get a proof that a parallel program respects the
+original formal semantics. In HPC, by applying this idea to parallel
+programming models like Legion, we can separate the application developer from
+having to prove correctness with respect to one or more increasingly complex,
+heterogeneous architechures. Instead, we can provide an intermediate layer,
+like Legion, that allows application developers to reason and prove correctness
+to much simpler operational semantics, and then prove that those semantics are
+retained even on the complex, heterogeneous architectures of todays
+supercomputers.
 
@@
 language for the CompCert project, Leroy et. al. found many cases where the
-informal specification was simply incomplete \cite{compcert}. Note that this is
-seperate notion from undefined behavior. Like unknown unknowns, gaps in a
-specification can make it impossible to prove desired properties of a system.
-In contrast, undefined behaviour is only a problem when programs exihibit it.
+informal specification was simply incomplete \cite{leroy2012compcert}. Note
+that this is seperate notion from undefined behavior. Like unknown unknowns,
+gaps in a specification can make it impossible to prove desired properties of a
+system.  In contrast, undefined behaviour is only a problem when programs
+exihibit it.
 
@@
 In the actual implementations this is handled both dynamically with futures
-and statically with dependency analysis \cite{regent}.
+and statically with dependency analysis \cite{slaughter2015regent}.
 
@@
 A third issue is that there is no way to dynamically create and destroy
-regions. In other words, it the semantics assumes static memory allocation.
+regions. In other words, if the semantics assumes static memory allocation.
 Of course, this differs from the actual implementation, and is an important,
@@
 Instead, it is embedded in an existing language. For the current implementations, 
-those languages are Regent and C/C++ \cite{regent, legion}. There are two 
-interesting paths towards addressing this issue. One would be to build a sort 
-of abstraction for languages that Legion is embedded in, and prove correctness
-with respect to that abstraction. This would likely be difficult to get right, as 
-capturing the important semantics of the languages that it would be embedded in
-would likely be a significant challenge. A second possibility for addressing this 
-issue would be choosing a particular semantics and focusing on that. In particular,
-combining the C API with the CompCert verified C work would likely be a
-fruitful, though challenging, venture.
+those languages are Regent and C/C++ \cite{slaughter2015regent}. There are two 
+interesting paths towards addressing this issue. One is to build a sort of
+abstraction for languages that Legion is embedded in, and prove correctness
+with respect to that abstraction. This would likely be difficult to get right,
+as capturing the important semantics of the languages that it would be embedded
+in would likely be a significant challenge. A second possibility for addressing
+this issue would be choosing a particular semantics and focusing on that. In
+particular, combining the C API with the CompCert verified C work would likely
+be a fruitful, though challenging, venture.
 
-All of these issues are addressable given enough time and effort. Some offer
-a nice opportunity for modularity within the tasking formalization. For example, 
+All of these issues are addressable given enough time and effort. Some offer a
+nice opportunity for modularity within the tasking formalization. For example,
 ordering tasks based on dependencies of values is common accross all tasking
-systems \cite{blumofe1995cilk, kaiser2014hpx}, and is likely a sort of sub-semantics that they could
-all share, reducing proof burden across different run-times.
+systems \cite{blumofe1995cilk, kaiser2014hpx}, and is likely a sort of
+sub-semantics that they could all share, reducing proof burden across different
+run-times.
 
@@
 Returning to the "P" in "HPC", one important current area of research is formal
-reasoning about performance \cite{performance}. One of the challenges for
-Legion, and runtime systems in general, is that the scheduler often has a
-non-trivial amount of work to do, leading to large overheads. In Legion, the
-combination of reasoning about permissions, memory locality, etc., makes this
-overhead particularly hard to predict.
+reasoning about performance \cite{mccarthy2017coq, chargueraud2015machine}. One
+of the challenges for Legion, and runtime systems in general, is that the
+scheduler often has a non-trivial amount of work to do, leading to large
+overheads. In Legion, the combination of reasoning about permissions, memory
+locality, etc., makes this overhead particularly hard to predict.
 
@@
 about \emph{any} property of a system, not just correctness. For example, there
-is the work by \cite{?} working towards formally verified complexity of code.
-Chlipala et. al. have also used techniques like those described here to construct
-and verify hardware description, where performance must be reasoned about down
-to single cycles \cite{braibant2013formal}.  This is one area where other tasking
-systems, like Cilk, have a better story than Legion. Blumofe et al. informally
-describe bounds on runtime overheads, which has been realized in practice as
-low overheads in the actual runtime \cite{blumofe1995cilk}.
+is the work by \cite{mccarthy2017coq} working towards formally verified
+complexity of code.  Chlipala et. al. have also used techniques like those
+described here to construct and verify hardware description, where performance
+must be reasoned about down to single cycles \cite{braibant2013formal}.  This
+is one area where other tasking systems, like Cilk, have a better story than
+Legion. Blumofe et al. informally describe bounds on runtime overheads, which
+has been realized in practice as low overheads in the actual runtime
+\cite{blumofe1995cilk}.
 

### 2017-08-26 correctnesshpc/paper.tex
change tone to be more formal

@@
 \begin{abstract}
-Legion is a programming model that uses logical regions of memory to reason about 
-locality and concurrency. We formalize the Legion semantics in a mechanized
-proof assistant, Coq, and show how this enables one to reason formally and
+The heterogenity and massive scale of extreme scale systems are making the
+traditional ad hoc implementation of programming models impractical. Formal
+specification and verification of programming models is a promising approach to
+ensuring correct behavior on heterogeneous systems at scales that would be
+impractical for empirical verification. In this paper we formalize the semantics
+of the Legion programming model which uses logical regions of memory to reason about 
+locality and concurrency in a mechanized
+proof assistant, Coq. We then show how this enables formal reasoning about the
+model and
 prove theorems about the semantics. We also discuss how to extend this work to
-prove that serial semantics are preserved and how given enough time, one could
-extend the effort to prove an implementation is correct.
+prove that serial semantics are preserved and how to extend the effort to
+provide a complete proof of the correctness of its implementation.
 \end{abstract}
@@
 
-\section{Intro}
-High performance computing, as the name implies, has long focused on
-performance. Expensive machines are purchased to run increasingly complex
-simulations, and every optimization can save huge amounts of money, both due to
-the initial cost of the machine and the cost of power to run it.
+\section{Introduction}
+As High Performance Computing (HPC) systems run increasingly complex simulations at
+increasingly large scales optimization of resource usage becomes proportionally
+more important in terms of both economic efficiency and scientific productivity. However the complexity of the interaction of the different
+factors being optimized, such as power consumption, computational intensity and
+I/O thoughput are outstripping the ability of individual researchers to understand or debug empirically.
 
-That said, what good is a fast computation if we aren't certain that the
-values it computes are correct? Presumably, these computations inform important
-decisions in areas ranging from climate to nuclear safety. We should strive to
-be as confident as possible that we aren't basing these decisions on buggy
-code.
+Given the importance of importance of HPC systems to issues of both national and
+economic security being able to efficiently and correctly specify, develop and
+validate these systems is an important research challenge with impacts 
+ranging from climate modeling to nuclear safety.
 
-The strongest guarantees of correctness come from the area of formal
-specifications and proofs. A specification is a description of how a program
-should behave, while proofs are used to show that an implementation conforms to
-a specification. \emph{Formal} specifications and proofs are defined as
-specifications and proofs that have been embedded in a machine-checked formal
+Formal specification and code proved by formal methods such as program logics
+provide the strongest guarantees of correctness. Specifications describe of how a program
+should behave, while proofs show that a implementations conform to
+their specifications. These \emph{formal} specifications and proofs 
+have been embedded in a machine-checked formal
 logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
@@
 machine checked logics has been shown to catch common errors in previously
-informal proofs, and therefore is often considered to provide the highest
-assurance of correctness currently possible \cite{clarke1996formal,
+informal proofs, and therefore provide higher 
+assurance of correctness than alternative methods \cite{clarke1996formal,
 yang2011finding}.
@@
 specifications enable formal correctness proofs for large software systems that
-were previously unattainable \cite{weng2016deepspec}. By integrating a formal
-specification tightly with an implementation, one can achieve the highest
-levels of confidence in code currently possible \cite{weng2016deepspec}. For
-example, the Compcert project implements an optimizing C compiler and proves it
-correct. The success of Compcert in removing compiler bugs has been well
-documented. We will discuss later how Compcert could be used in building
+were previously thought unattainable \cite{weng2016deepspec}. By integrating a formal
+specification tightly with an implementation, one can achieve the high
+levels of confidence in a codes' adherence to its
+specification\cite{weng2016deepspec}. For example, the Compcert project
+provides a specification for the C language, implements it as series of
+composed modules which are proved correct independently before being combined
+into an optimizing C compiler. The success of Compcert in removing compiler bugs has been well
+documented ~\cite{leroy2012compcert}. We will discuss later how Compcert could be used in building
 verifiable HPC systems.
 
-In HPC, programming models have become an increasingly important way to ensure
-that code stays maintainable, performant, and portable \cite{diaz2012survey}.
-We claim that these programming models provide a good target for formalization:
-any correctness arguments can be leveraged by every program using the model. It
-is exactly this kind of modularization of proofs that has been so successful in
-proving programs correct \cite{weng2016deepspec}.
+HPC programming models are particularly attractive target for these composable
+specifications and verified implementations because of their importance 
+ensuring that simulation codes stay maintainable, performant, and portable at
+extreme scales\cite{diaz2012survey}.
+We claim in particular that formalizing programming models have wide ranging
+impact across the spectrum of HPC application development --any composable correctness
+proofs can be leveraged into proofs of correctness of every program using the
+model rather than having to be reimplemented on a per application basis.
 
-For this paper, we take a recent programming model, Legion, and formalize its
-semantics and type system. We then prove a few supporting lemmas as examples of
+In this paper, we apply these composable specification and verification methods
+to a recent programming model, Legion, by formalize its
+semantics and type system in the Coq proof assistant. We then prove a few supporting lemmas as examples of
 the kinds of things one can prove in an HPC settings. Finally, we discuss the
 some of the possible paths towards a full, formally correct implementation.
-This is no small task, as we discuss later, and the ways forward are numerous
-and difficult. But if we believe that HPC informs important decisions, we must
-rise to the challenge.
+
+This paper makes the following contributions:
+\begin{itemize}
+  \item An analysis of the requirements for specifying and verifying an HPC
+    programming model. 
+  \item A formalization of the semantics of the Legion Programming Model in the
+    Coq proof assistant.
+  \item A description of a path to a fully verified programming model with provable
+    correctness properties using soundness and completeness proofs.
+  \item An exploration of the challenges to formally specifying the Legion runtime.
+\end{itemize}
 
@@
 
-Legion is a programming model for high performance computing that enables
+Legion is an HPC programming model that enables
 reasoning about locality and concurrency via logical regions of memory. It has
@@
 burden of reasoning about the non-determinism that arises from concurrent
-computation. For this paper, we'll be formalizating work on from Triecher et.
-al \cite{treichler2013language}. While there has been much work extending
-Legion since then, the core semantics has stayed relatively stable
+computation. Specifically we'll be formalizating work on from Triecher et.
+al \cite{treichler2013language}. While there has been subsequent work extending
+Legion's capabilities, the core semantics of the Programming Model has stayed relatively stable
 \cite{slaughter2015regent, treichler2016dependent}.
@@
 Legion is able to ensure that certain code won't interfere or alias, and can
-therefore be safely, at least with respect to those memory regions, run in
-parallel. This theory has been put into practice to impressive effect, and has
-been shown to scale both within a node and between \cite{bauer2012legion,
-treichler2014realm}. That said, there are still subtle bugs that occasionally
-show up in the implementation. One goal of this paper is to move towards
+therefore be safely --at least with respect to those memory regions-- run in
+parallel. This partitioning mechanism has been shown to be effective in enabling
+applications to scale both on and off node~\cite{bauer2012legion,
+treichler2014realm}. However subtle bugs still occur 
+ in the implementation. One goal of this paper is to move towards
 formally and systematically removing \emph{all} bugs from the programming
-model, giving everyone who uses is far more certainty that their code is
-correct.
+model, giving application developers and researcher quantifiable certainty that
+their code correctly implements the semantics of the science being simulated.
 
-To move towards this goal, we turn to mechanized proof assistants.
+Mechanized proof assistants provide one tool to reach this certainty.
 Improvements to proof assistants over recent decades have made them a powerful
-tool for implementing and reasoning about code. From provably correct compilers
-to provably correct operating systems, proof assistants enable extremely high
-certainty in correctness even for large scale software projects.
+tool for implementing and reasoning about code. Recent research using proof
+assistants has been able to verify the correctness of compilers~\cite{leroy2012compcert} and 
+and operating systems~\cite{gu2011certikos}.
 
-Coq, in particular, has proven itself to be a particularly potent dependently
-typed proof assistant. At its core is the Calculus of Inductive Constructions,
-a small dependently typed programming language and logic, capable of powerful
-abstractions for formal proofs \cite{bertot2013interactive}. In the last 10
-years great strides have been made in using it to both build systems and prove
+In particular, the Coq proof assistant has been effective for a variety of
+verification tasks. Coq uses the Calculus of Inductive Constructions as its type 
+theory allowing it to implement
+a small dependently typed programming language and logic, capable of  
+formal proofs of arbitrary mathematical objects\cite{bertot2013interactive}. In the last 10
+years it has been used to both build systems via program extraction and prove
 useful properties about them \cite{leroy2012compcert, anand2017certicoq,
@@
 
-Specifications come in a wide range of formats. Many, like the OpenMP
-specification, are very informal, primarily consisting of english descriptions
+To implement a specification of a Programming Model in a proof assistant like
+Coq it is customary to specify that model as a set of inference rules describing
+the typing rules (the 'Statics') which describe valid terms in the model and its
+operational semantics (the 'Dynamics') which describe the valid state
+transitions of the model. However, most Programming Models are not specified at
+this level of detail. Many, like the OpenMP
+specification, are very informal, primarily consisting of English descriptions
 of how different constructs should behave \cite{dagum1998openmp}. Others, like
 Legion's, are written by hand in the using standard notation for logical
-relations. For this paper, when we refer to a \emph{formal} specification, we
-are referring to a specification defined in a formal logic, eg. Coq.
+relations for types and operational semantics with informal proofs based on these
+rules. These approaches are contrasted to \emph{formal} specifications, 
+which are specifications defined in a machine checkable formal logic, eg. Coq.
 
-While semantics defined by hand as logical relations are a huge improvement
-over natural language specifications, there is still much room for improvement.
-Like type checking a program, type checking a proof exposes many "bugs" in a
-proof. This is not to say, of course, that formal specifications and proofs are
-infallible. On the contrary, it is quite easy to get a specification wrong.
-That said, often, if a there is a mistake in a spec, the process of proving it
-formally will expose that mistake, allowing for its correction.
+The machine checkable proofs provided by these formal specifications make it
+possible to encode both the natural language and logical specifications in a
+formal language and verify the correctness in a manner analogous to type checking a program, exposing "bugs" in 
+proofs and incompleteness in specifications. It is worth nothing, however, that
+these that formal specifications and proofs implemented in proof assistant are
+not infallible. Mistakes in specification can lead to incorrect assumptions and
+behavior and the process of recognizing and correcting these inconsistencies is
+an important part of the proving process.
 
@@
 
-A program without a specification cannot be incorrect, it can only be
-surprising \footnote{Paraphrase of J.J. Horning, 1982}. One of the most
+Experience in the specification of programs and fixing bugs in those
+specification has led to the aphorism that ``program without a specification cannot be incorrect, it can only be
+surprising''\footnote{Paraphrase of J.J. Horning, 1982}. Its insight is that one
+of the most
 important and challenging aspects of proving correctness is defining a formal
-specification of how a program should behave. Indeed, there is currently a
-large NSF effort involving many of the best minds in programming languages
-dedicated to improving our ability to do just this \cite{weng2016deepspec}. The choice
-of specification has far-reaching consequences in how easy it is to later prove
+specification of program behavior. Developing correct and modular specifications
+is a subject of active research including the NSF DeepSpec
+project\cite{deepspecnsf} which is exploring abstraction and composition
+mechanisms for program specification. Particularly important in the development
+of specifications is that the assumptions stated in a 
+specification have far-reaching consequences in how easy it is to later prove
 correctness properties.
 
-One thing that makes Legion unique among high performance runtime systems is
-that it provides a specification of its operational semantics \cite{treichler2013language}.
+The Legion Programming Model provides an attractive target for formal
+specification and verification because  --unique among high performance runtime systems-- Legion
+provides a specification of its operational semantics \cite{treichler2013language}.
 We take this specification of and formalize it in Coq, enabling us to prove
-lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
-implementation of the syntax, and Figure~\ref{semantics} for semantics. All
+lemmas and theorems about it. Figure~\ref{syntax} provides a Coq
+implementation of the syntax, and Figure~\ref{semantics} the semantics. All
 code is available online at \texttt{https://github.com/lanl/formallegion}.
@@
 parallel execution. This is in contrast to programming models like Cilk, which
-claim a weaker property: a serial execution is one of many valid executions.
+claim a weaker property: a serial execution is one of many valid possible executions.
 
@@
 While so far we have discussed the formalization of the \emph{specification},
-we have largely left discussion of proofs by the wayside. In this section, we
+we have largely left proofs undiscussed. In this section, we
 discuss the challenges of proving useful properties of real world systems, and
@@
 important construction are mappers, \texttt{M}, which map logical regions
-\texttt{r} with a set of physical locations. We include some definitions as
+\texttt{r} to a set of physical locations. We include some definitions as
 well, notably the notion that a mapping respects (\texttt{resp}) a set of
@@
 
-One important property when proving correctness in Legion is to show that the
-relations are equivalent to their closure, with respect to what they imply
-about locations in those regions.  Figure~\ref{regionrelproof} for the
-statement of this proof. Intuitively, this means that if a region is a
+One important property to show when proving correctness of programs written the
+Legion model is to show that relations are equivalent to their closure (ie. that
+composed relations are equivalent to relations in isolation), especially with
+respect memory locality. Figure~\ref{regionrelproof} provides a proof for this
+assertion. Intuitively, this means that if a region is a
 subregion of another region, and that region is a subregion of a third, then
@@
 
-To do this we are writing our proofs of progress and preservation (ie. that well
+To do this we are writing our proofs of completeness and soundness (ie. that well
 typed expressions evaluate to a value or diverge and that every transition in
@@
 
-The theorems we are using are described in Figure~\ref{preservationtheorem} and
-Figure~\ref{progresstheorem} (see \texttt{https://github.com/lanl/formallegion}
-for the in progress proofs).
+The theorems we are using are described in Figure~\ref{soundnesstheorem} and
+Figure~\ref{completenesstheorem} (see \texttt{https://github.com/lanl/formallegion}
+for Coq versions of the proofs).
 
@@
 
-The work presented in this paper is a small fraction of a small fraction of the
+The work presented in this paper is a small fraction of the
 amount of work required to build a complete, correct-by-construction HPC
 application. In this section, we outline some of the possible paths towards
-this worthy goal.
+this goal.
 
-As mentioned earlier in the paper, one of the keystones for large formally
+As mentioned earlier, one of the important characteristics of large formally
 correct systems is modularity. This modularity has two purposes. On the one
-hand, it enables ease of implementation. The second, analagous advantage, is
-that in a similar way modularity eases the burden of proofs. By defining
-specifications as abstractions, one can separate concerns of proofs. For
+hand, it enables ease of implementation. The second, analogous advantage, is
+that modularity eases the burden of proofs. By defining
+specifications as abstractions, one can separate proof concerns into proofs of
+individual components or modules. For
 example, by composing a proof that a program's serial semantics comply with a
 formal specification with a proof that a parallel language respects those
-serial semantics, we can get a proof that a parallel program respects the
+serial semantics, it is possible derive a proof that a parallel program respects the
 original formal semantics. In HPC, by applying this idea to parallel
@@
 having to prove correctness with respect to one or more increasingly complex,
-heterogeneous architechures. Instead, we can provide an intermediate layer,
+heterogeneous architectures. Instead, we can provide an intermediate layer,
 like Legion, that allows application developers to reason and prove correctness
 to much simpler operational semantics, and then prove that those semantics are
-retained even on the complex, heterogeneous architectures of todays
-supercomputers.
+retained even on the complex, heterogeneous architectures of today's
+HPC Systems.
 
-Of course this is much easier said than done, and we address some of the challenges
-faced by such an effort in the following subsections.
+However modularly specifying HPC is not easy and is a challenging research
+problem. We illustrate some of these difficulties in the following section.
 
@@
 informal specification was simply incomplete \cite{leroy2012compcert}. Note
-that this is seperate notion from undefined behavior. Like unknown unknowns,
+that this is separate notion from undefined behavior. Like unknown unknowns,
 gaps in a specification can make it impossible to prove desired properties of a
-system.  In contrast, undefined behaviour is only a problem when programs
+system.  In contrast, undefined behavior is only a problem when programs
 exihibit it.
 
-In formalizing Legion, we discovered that while there is an impressive theory
+In formalizing Legion, we have discovered that while there is a rich theory
 ensuring that the side effects of concurrent evaluations preserve serial
@@
 \texttt{write}, despite the fact that the value being written, \texttt{x},
-generally will need to be read before being written. This isn't catastrophic
-for the semantics, as the returned value and resulting heap are equivalent, 
-as desired, but it limits the ability to reason effectively about concurrency. 
+generally will need to be read before being written. This inconsistency isn't catastrophic
+for the semantics, the returned value and resulting heap are equivalent. 
+ However it limits the ability to reason effectively about concurrency. 
 In the actual implementations this is handled both dynamically with futures
@@
 Of course, this differs from the actual implementation, and is an important,
-though difficult, semantics to formalize. As memory heirarchies get more 
+though difficult, semantics to formalize. As memory hierarchies get more 
 sophisticated, the need for tools for reasoning about memory use and locality
@@
 
-Returning to the "P" in "HPC", one important current area of research is formal
+In addition to semantic correctness, one important current area of research in
+formal methods is formal
 reasoning about performance \cite{mccarthy2017coq, chargueraud2015machine}. One
@@
 must be reasoned about down to single cycles \cite{braibant2013formal}.  This
-is one area where other tasking systems, like Cilk, have a better story than
+is one area where other tasking systems, like Cilk, are better understood than
 Legion. Blumofe et al. informally describe bounds on runtime overheads, which
@@
 
-Ideally, we'd like to make similar claims about Legion runtime overheads.
-Given a formal model of the work the runtime system has to do, one can reason
+Ideally, it would be possible to make similar claims about Legion runtime overheads.
+Given a formal model of the work done by the runtime system, one can reason
 about bounds on overhead. Currently, due to the complexity of the runtime
@@
 
-While we've focused on Legion, there are other parallel programming systems
-that offer exciting opportunities for formalization. While some (looking at
-you, OpenMP), are likely much to large and complex to formalize without
-herculean effort, others have simple enough semantics that formalization is
-likely within reach. One in particular that is attractive due to its simplicity
-is the recent work on parallel extensions to LLVM, the Tapir project
-\cite{schardl2017tapir}. Building on existing formalization efforts for LLVM
+Although the focus of this paper has been on the Legion Programming Model, there are other parallel programming systems
+that offer exciting opportunities for formalization. While some, such as
+ OpenMP, are likely much to large and complex to formalize without
+significant research effort, others have simple enough semantics that formalization is
+likely tractable.the Tapir project
+\cite{schardl2017tapir} is a particularly attractive research target due to its simplicity
+and the recent work on parallel extensions to LLVM. Building on existing formalization efforts for LLVM
 \cite{zhao2012formalizing}, extending existing semantics with the Tapir instructions for
-concurrency should be viable. Again, there would be many challenges,
-particularly in the area of formal reasoning about memory models, but the
-approach should be viable.
+concurrency should be achievable in reasonable time despite challenges in
+specifying its memory model.
 
-As mentioned earlier, there is also the important notion of \emph{shared
+As mentioned earlier, there is also the important concept of \emph{shared
 semantics accross different tasking systems}. This notion of shared semantics
@@
 
-We hope to have convinced the reader of the importance and viability of proving
-parallel programming systems in HPC correct. While we've only taken small steps
-on the path to provably correct Legion, our hope is that it will inspire
-support for such ventures by providing a convincing story on how to move
-forward. There is ever- increasing evidence that this class of deep
-specifications and certified software is an incredibly powerful tool for
-proving desirable properties about our large software systems. We'd like to see
-more focus on correctness in HPC, and we think deep specifications and
-certified proofs are the way forward.
+In this paper, we have proposed a path to formally specifying HPC programming
+models using mechanized proof assistants. To evaluate this proposal we have given a formal
+semantics of the Legion Programming Model and proved its correctness regarding
+memory interleavings between relations. We have also described mechanisms to
+prove the soundness and completeness of our approach as well as the potential
+difficulties in dealing with potential gaps in specification. Finally we
+concluded by proposing the Tapir system as a viable next step in the
+specification of HPC programming models.
 

### 2017-08-26 correctnesshpc/paper.tex
fix margins on figure

@@
 \begin{BVerbatim}
-Theorem preservation : forall (e5:e) (t:T) (E:E) (M:M) (L:L) 
-(H:H) (S:list (e * e)) (C:list e) (T:T),
+Theorem preservation : forall (e5:e) (t:T) (E:E) (M:M) 
+(L:L) (H:H) (S:list (e * e)) (C:list e) (T:T),
     exists v,
       eval M L H S C e5 v E ->
-      typing EmptyContext EmptyPriv EmptyConstraint e5 T  ->
+      typing NilContext NilPriv NilConstraint e5 T  ->
       exists G P Omega T, typing G P Omega v T.
 \end{BVerbatim}
-\caption{preservation theorem}
-\label{preservationtheorem}
+\caption{soundness theorem}
+\label{soundnesstheorem}
 \end{figure}
@@
 (C:list e) (T1 T2:T) L (S:list (e * e)),
-    typing EmptyContext EmptyPriv EmptyConstraint e5 T1  ->
+    typing NilContext NilPriv NilConstraint e5 T1  ->
     (exists E v ,
@@
 \end{BVerbatim}
-\caption{progress theorem}
-\label{progresstheorem}
+\caption{completeness theorem}
+\label{completenesstheorem}
 \end{figure}

### 2017-08-26 correctnesshpc/paper.tex
George edits 2.1

@@
 ensuring correct behavior on heterogeneous systems at scales that would be
-impractical for empirical verification. In this paper we formalize the semantics
-of the Legion programming model which uses logical regions of memory to reason about 
-locality and concurrency in a mechanized
-proof assistant, Coq. We then show how this enables formal reasoning about the
-model and
-prove theorems about the semantics. We also discuss how to extend this work to
-prove that serial semantics are preserved and how to extend the effort to
-provide a complete proof of the correctness of its implementation.
+impractical for empirical verification. In this paper we formalize the
+semantics of the Legion programming model which uses logical regions of memory
+to reason about locality and concurrency in a mechanized proof assistant, Coq.
+We then show how this enables formal reasoning about the model and prove
+theorems about the semantics. We also discuss how to extend this work to prove
+that serial semantics are preserved and how to extend the effort to provide a
+complete proof of the correctness of its implementation.
 \end{abstract}
@@
 \section{Introduction}
-As High Performance Computing (HPC) systems run increasingly complex simulations at
-increasingly large scales optimization of resource usage becomes proportionally
-more important in terms of both economic efficiency and scientific productivity. However the complexity of the interaction of the different
-factors being optimized, such as power consumption, computational intensity and
-I/O thoughput are outstripping the ability of individual researchers to understand or debug empirically.
+As High Performance Computing (HPC) systems run increasingly complex
+simulations at increasingly large scales optimization of resource usage becomes
+proportionally more important in terms of both economic efficiency and
+scientific productivity. However the complexity of the interaction of the
+different factors being optimized, such as power consumption, computational
+intensity and I/O thoughput are outstripping the ability of individual
+researchers to understand or debug empirically.
 
-Given the importance of importance of HPC systems to issues of both national and
-economic security being able to efficiently and correctly specify, develop and
-validate these systems is an important research challenge with impacts 
-ranging from climate modeling to nuclear safety.
+Given the importance of importance of HPC systems to issues of both national
+and economic security, being able to efficiently and correctly specify,
+develop, and validate these systems is an important research challenge with
+impacts ranging from climate modeling to nuclear safety.
 
 Formal specification and code proved by formal methods such as program logics
-provide the strongest guarantees of correctness. Specifications describe of how a program
-should behave, while proofs show that a implementations conform to
-their specifications. These \emph{formal} specifications and proofs 
-have been embedded in a machine-checked formal
-logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
+provide the strongest guarantees of correctness. Specifications describe how
+a program should behave, while proofs show that a implementations conform to
+their specifications. \emph{Formal} specifications and proofs are those that
+have been embedded in a machine-checked formal logic
+\cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
 jackson1994nuprl}. This process of formalizing specifications and proofs in
 machine checked logics has been shown to catch common errors in previously
-informal proofs, and therefore provide higher 
-assurance of correctness than alternative methods \cite{clarke1996formal,
-yang2011finding}.
+informal proofs, and therefore provide higher assurance of correctness than
+alternative methods \cite{clarke1996formal, yang2011finding}.
 
@@
 specifications enable formal correctness proofs for large software systems that
-were previously thought unattainable \cite{weng2016deepspec}. By integrating a formal
-specification tightly with an implementation, one can achieve the high
+were previously thought unattainable \cite{weng2016deepspec}. By integrating a
+formal specification tightly with an implementation, one can achieve the high
 levels of confidence in a codes' adherence to its
@@
 composed modules which are proved correct independently before being combined
-into an optimizing C compiler. The success of Compcert in removing compiler bugs has been well
-documented ~\cite{leroy2012compcert}. We will discuss later how Compcert could be used in building
-verifiable HPC systems.
+into an optimizing C compiler. The success of Compcert in removing compiler
+bugs has been well documented ~\cite{leroy2012compcert}. We will discuss later
+how Compcert could be used in building verifiable HPC systems.
 
@@
 ensuring that simulation codes stay maintainable, performant, and portable at
-extreme scales\cite{diaz2012survey}.
-We claim in particular that formalizing programming models have wide ranging
-impact across the spectrum of HPC application development --any composable correctness
-proofs can be leveraged into proofs of correctness of every program using the
-model rather than having to be reimplemented on a per application basis.
+extreme scales\cite{diaz2012survey}.  We claim in particular that formalizing
+programming models have wide ranging impact across the spectrum of HPC
+application development --any composable correctness proofs can be leveraged
+into proofs of correctness of every program using the model rather than having
+to be reimplemented on a per application basis.
 
 In this paper, we apply these composable specification and verification methods
-to a recent programming model, Legion, by formalize its
-semantics and type system in the Coq proof assistant. We then prove a few supporting lemmas as examples of
-the kinds of things one can prove in an HPC settings. Finally, we discuss the
-some of the possible paths towards a full, formally correct implementation.
+to a recent programming model, Legion, by formalize its semantics and type
+system in the Coq proof assistant. We then prove a few supporting lemmas as
+examples of the kinds of things one can prove in an HPC settings. Finally, we
+discuss the some of the possible paths towards a full, formally correct
+implementation.
 
@@
 The theorems we are using are described in Figure~\ref{soundnesstheorem} and
-Figure~\ref{completenesstheorem} (see \texttt{https://github.com/lanl/formallegion}
+Figure~\ref{preservationtheorem} (see \texttt{https://github.com/lanl/formallegion}
 for Coq versions of the proofs).
@@
 \begin{BVerbatim}
-Theorem preservation : forall (e5:e) (t:T) (E:E) (M:M) 
-(L:L) (H:H) (S:list (e * e)) (C:list e) (T:T),
-    exists v,
-      eval M L H S C e5 v E ->
-      typing NilContext NilPriv NilConstraint e5 T  ->
-      exists G P Omega T, typing G P Omega v T.
+Theorem preservation : ∀ e t E M L H S C T,
+  ∃ v,
+    eval M L H S C e v E →
+    typing [] [] [] e T → 
+    ∃ G P Ω, typing G P Ω v T.
 \end{BVerbatim}
-\caption{soundness theorem}
-\label{soundnesstheorem}
+\caption{Preservation theorem}
+\label{preservationtheorem}
 \end{figure}
@@
 \begin{BVerbatim}
-Theorem progress : forall  (e5:e) (t:T) (M:M) (H:H)  
-(C:list e) (T1 T2:T) L (S:list (e * e)),
-    typing NilContext NilPriv NilConstraint e5 T1  ->
-    (exists E v ,
-        is_value v ->
-        eval M L H S C e5 v E).
+Theorem soundness : ∀ e t M H C T1 T2 L S,
+  typing [] [] [] e T1  →
+  (∃ E v,
+    is_value v →
+    eval M L H S C e v E).
 \end{BVerbatim}
-\caption{completeness theorem}
-\label{completenesstheorem}
+\caption{Soundness theorem}
+\label{soundnesstheorem}
 \end{figure}

### 2017-08-26 correctnesshpc/paper.tex

@@
 
-The theorems we are using are described in Figure~\ref{soundnesstheorem} and
-Figure~\ref{preservationtheorem} (see \texttt{https://github.com/lanl/formallegion}
+The theorems we are using are described in Figure~\ref{preservationtheorem} and
+Figure~\ref{soundnesstheorem} (see \texttt{https://github.com/lanl/formallegion}
 for Coq versions of the proofs).
@@
 \begin{BVerbatim}
-Theorem soundness : ∀ e t M H C T1 T2 L S,
+Theorem Soundness : ∀ e t M H C T1 T2 L S,
   typing [] [] [] e T1  →
@@
 
-The work presented in this paper is a small fraction of the
-amount of work required to build a complete, correct-by-construction HPC
-application. In this section, we outline some of the possible paths towards
-this goal.
+The work presented in this paper is a small fraction of the amount of work
+required to build a complete, correct-by-construction HPC application. In this
+section, we outline some of the possible paths towards this goal.
 
@@
 specifications as abstractions, one can separate proof concerns into proofs of
-individual components or modules. For
-example, by composing a proof that a program's serial semantics comply with a
-formal specification with a proof that a parallel language respects those
-serial semantics, it is possible derive a proof that a parallel program respects the
-original formal semantics. In HPC, by applying this idea to parallel
-programming models like Legion, we can separate the application developer from
-having to prove correctness with respect to one or more increasingly complex,
-heterogeneous architectures. Instead, we can provide an intermediate layer,
-like Legion, that allows application developers to reason and prove correctness
-to much simpler operational semantics, and then prove that those semantics are
-retained even on the complex, heterogeneous architectures of today's
-HPC Systems.
+individual components or modules. For example, by composing a proof that a
+program's serial semantics comply with a formal specification with a proof that
+a parallel language respects those serial semantics, it is possible derive a
+proof that a parallel program respects the original formal semantics. In HPC,
+by applying this idea to parallel programming models like Legion, we can
+separate the application developer from having to prove correctness with
+respect to one or more increasingly complex, heterogeneous architectures.
+Instead, we can provide an intermediate layer, like Legion, that allows
+application developers to reason and prove correctness to much simpler
+operational semantics, and then prove that those semantics are retained even on
+the complex, heterogeneous architectures of today's HPC Systems.
 
